Syntax Is All You Need: A Universal-Language Approach to Mutant Generation

### 2023-03-29 paper/main.tex
some edits, comments in intro

Mutation testing, though introduced in the late
1970s~\cite{demillo1978hints,mathur2012foundations,demillo1978hints},
has largely been a topic of academic interest rather than widespread
practice until quite recently.  However, with the adoption of mutation
testing by bellwether software industry companies including
Google~\cite{GoogleMut}, Meta~\cite{BellerFacebookMutation}, and
Amazon~\cite{AmazonMut}, interest in using mutation testing for
real-world projects has grown widely.

Interest in mutation testing has become widespread enough in the open
source community that a popular (more than 300 stars on GitHub)
repository, ``Awesome Mutation Testing,'' lists more than 40 tools,
covering almost twenty languages:
\url{https://github.com/theofidry/awesome-mutation-testing}.  Some of
these tools are the products of academic research, but many are
essentially the hobby projects of developers interested in enabling
mutation testing for their favorite language.

However, the same website demonstrates the existence of a sad history
of mutation testing tools: mutation tools have a tendency to be
developed for a while and then abandoned.   The page listing abandoned
mutation testing tools
(\url{https://github.com/theofidry/awesome-mutation-testing/blob/master/abandoned.md})
lists almost half as many tools as are under active maintenance, and
these tools cover approximately half as many languages.  Each of these
tools required considerable development time and effort, and was
popular enough at some point to be worth noting in the ``Awesome
Mutation Testing'' resource.  Many of the abandoned tools have GitHub
repositories (some archived) with more than 20 stars, and so may well
have had users who at some point faced the reality of no future bug
fixes or changes to address changes to the programming language being
mutated\footnote{While this is a minor concern for languages such as C
  that are relatively stable, it means that a mutation tool may fail
  to parse C++ 2020, and is particularly problematic for somewhat
  fast-moving languages such as Rust.}.  There
appears to be a significant risk in relying on any mutation tools other
than the most widely used and well-supported, such as PIT.

Moreover, newer languages, including ones with considerable interest
in the development community, often lack mutation testing tools.  Even
languages with a long history may long lack a working mutation testing
tool; e.g., it was not until 2014 that Haskell had such a tool
\cite{mucheck}.  Furthermore, that tool is no longer
supported\footnote{The last commits or package updates were in 2015, and the
primary author confirms that changes to {\tt haskell-src-ext} since
introduced render it unusable at present.}, and it
appears that no successor tool has appeared in the ensuing
years\footnote{There is FitSpec~\cite{FitSpec}, which is maintained,
  but FitSpec does not perform traditional mutation testing or produce
  source code (or bytecode) mutants; it rather
  performs black-box type-aware mutation of function return values.}.
Haskell developers (until now, see below) have no option for
performing true mutation testing.

A great deal of research effort has been devoted to devising clever
schemes for reducing the computational burden of mutation testing over
the years; in practice, however, the recent adoption of mutation
testing arguably relies more on the availability of vast (cloud) computing
resources than any particular approach from the literature, beyond
heuristic restrictions on the mutants used (e.g., one per line, or
only some siimple operators).   On the other hand, almost no
research has addressed perhaps the most central issue of
mutation testing: in the absence of a working mutation testing
tool for the language in which one is developing software, one cannot
obtain the benefits of mutation testing.  The absence of such tools is
due to the high human cost of developing and maintaining mutation
testing tools.  Ideally, mutation testing would be available for
\emph{any} newly proposed programming language, as soon as that
language has any community of users at all.  In the present
state-of-the-art, however, it may be years before even rudimentary
tools appear, and they may ``disappear'' shortly there-after.

>>>>

Mutation testing, though introduced in the late
1970s~\cite{demillo1978hints,mathur2012foundations,demillo1978hints},
has long been a topic of academic interest.  With the recent adoption of mutation
testing by bellwether software industry companies including
Google~\cite{GoogleMut}, Meta~\cite{BellerFacebookMutation}, and
Amazon~\cite{AmazonMut}, interest in using mutation testing for
real-world projects has grown widely in practice. 
%
Indeed, interest in mutation testing has become widespread enough in the open
source community that a popular (more than 300 stars on GitHub)
repository, ``Awesome Mutation Testing,'' lists more than 40 tools,
covering almost twenty languages:
\url{https://github.com/theofidry/awesome-mutation-testing}.  Some of
these tools are the products of academic research; others are
essentially the hobby projects of developers interested in enabling
mutation testing for their favorite language.

This diversity of tools highlights the growth of widespread interest in mutation
testing.  However, it also demonstrates various ways that bringing mutation testing to
typically polyglot modern projects is cumbersome. Each language (or IR, at
minimum) requires its own underlying mutation and testing framework.
\clg{Alex, can something be said about 
  how many LOC the tools we look at are on average, this early?}

Newer languages, including ones with considerable interest
in the development community, often lack mutation testing tools.  Even
languages with a long history may long lack a working mutation testing
tool; e.g., it was not until 2014 that Haskell had such a tool~\cite{mucheck}. 
%
Furthermore, the Haskell mutation tool is no longer
supported,\footnote{The last commits or package updates were in 2015, and the
primary author confirms that changes to {\tt haskell-src-ext} since
introduced render it unusable at present.} and it
appears that no successor tool has appeared in the ensuing
years\footnote{There is FitSpec~\cite{FitSpec}, which is maintained,
  but FitSpec does not perform traditional mutation testing or produce
  source code (or bytecode) mutants; it rather
  performs black-box type-aware mutation of function return values.}.
Haskell developers (until now, see below) have no option for
performing true mutation testing.

Indeed, abandonment is a feature of the sad history
of many mutation testing tools, exacerbating the challenges of its uptake.  The page listing abandoned
mutation testing tools
(\url{https://github.com/theofidry/awesome-mutation-testing/blob/master/abandoned.md})
lists almost half as many tools as are under active maintenance. Each tool represents 
considerable development effort, and was
popular enough to be noted in the ``Awesome
Mutation Testing'' resource. Many of the abandoned repositories have more than
20 stars, suggesting inconvenienced users, especially for tools targeting
evolving languages. While this is a minor concern for languages such as C
that are relatively stable, it means that a mutation tool may fail
to parse C++ 2020, and is particularly problematic for 
faster-moving languages like Rust.

%There
%appears to be a significant risk in relying on any mutation tools other
%than the most widely used and well-supported, such as PIT.
A great deal of research effort has been devoted to devising clever
schemes for reducing the computational burden of mutation testing over
the years;\clg{a few references?} in practice, however, the recent adoption of mutation
testing arguably relies more on the availability of vast (cloud) computing
resources than any particular approach from the literature, beyond
heuristic restrictions on the mutants used (e.g., one per line, or
only some siimple operators).   On the other hand, almost no
research has addressed perhaps the most central issue of
mutation testing: in the absence of a working mutation testing
tool for the language in which one is developing software, one cannot
obtain the benefits of mutation testing.  The absence of such tools is
due to the high human cost of developing and maintaining mutation
\clg{OK I don't mind this argument, BUT it kind of seems in conflict with the
  argument above that says there are a lot of tools and mutation testing is
  popular.  I think it might be worth shifting the emphasis of this paragraph to
  be to motivate source-level mutation testing over IR, because that's missing
  so far and would be good to motivate.  Note that it's also an argument made in
  the abstract, which I think is a particularly good summary of the paper.}
testing tools.  Ideally, mutation testing would be available for
\emph{any} newly proposed programming language, as soon as that
language has any community of users at all.  In the present
state-of-the-art, however, it may be years before even rudimentary
tools appear, and they may ``disappear'' shortly there-after.
